\documentclass[11pt]{article}
\input{preamble.tex}

\begin{document}
\maketitle

Source code for this assignment can be found on \href{https://github.com/laspp/HPC/blob/main/labs/02-Assignment1-OpenMP/Assignment1.md}{GitHub}.
Our scripts and modified source code can be found in our \href{https://github.com/Edelwy/HPC}{\textbf{GitHub} repository}. \href{https://en.wikipedia.org/wiki/Seam_carving}{\textbf{Seam carving}} resizes an image by repeatedly finding a low-energy vertical seam, removing one pixel per row, and shifting the remaining pixels. Each outer iteration applies three phases in order: we compute a the Sobel energy map, find the cheapest seam by dynamic programming and trace the cheapest seam from the top row down.

\section{Implementation}

The baseline parallel version uses \href{https://www.openmp.org/wp-content/uploads/OpenMP-4.0-C.pdf}{\textbf{OpenMP}} library and can be enabled via \texttt{USE\_OMP} macro. This includes parallization per pixel in the energy calculation, parallelization per row in seam identification and parallelization in pixel removal per seam partitions which are dependant on height.  

%TODO 

An \textbf{optimization} is supported via \texttt{USE\_DYNAMIC\_THREADS} macro, which keeps the same pragmas but sets \textbf{different number of threads} at the boundaries between phases so each stage can use a different size. The heuristic dynamically selects the number of threads based on the amount of available work in each phase of the algorithm. For the energy computation, the thread count scales with the total workload divided by a tuning constant \texttt{THREADS\_ENERGY\_CONSTANT (C1)}, allowing utilization of cores for large images while avoiding excessive overhead for small ones.The thread count for the dynamic programming phase is derived from the width and \texttt{THREADS\_DP\_CONSTANT (C2)} and capped at half of the available threads to reduce synchronization overhead. Similarly, the seam removal phase is memory-bound and parallelized across image rows, so its thread count is based on the image height and \texttt{THREADS\_REMOVAL\_CONSTANT (C3)} and also conservatively limited to prevent contention for memory bandwidth. The constants were set to different valaues and tested to find the right balance as can be seen in results.

We also implemented \textbf{greedy seam removal} which removes multiple seams without calculating the energy for each removal and compared the results by looking at speed up and picture quality. In order to achieve further paralelization we divided the picture into blocks where width is divided by the number of threads in use. The path finding is limited to the block so that these can be done in parallel independently. This is supported via \texttt{USE\_GREEDY\_REMOVAL} macro.

\section{Comparing Parallelization Without Optimization}

The following table lists the average \textbf{total time in seconds} used for the sequential and parallel algorithm implementation over 5 repetitions and run on different number of threads. One thread data is included in order to test library overhead when comparing to pure sequential implementation.

\begin{figure}[H]
    \begin{minipage}{\textwidth}
        \begin{table}[H]
            \centering
            \begin{tabular}{|l|*{7}{p{1.25cm}|}} \hline
                \cellcolor{blue!20}\textbf{image / threads} & \cellcolor{blue!20}\texttt{seq.} & \cellcolor{blue!20}\texttt{1} & \cellcolor{blue!20}\texttt{2} & \cellcolor{blue!20}\texttt{4} & \cellcolor{blue!20}\texttt{8} & \cellcolor{blue!20}\texttt{16} \\ \hline
                \texttt{720$\times$480} & 0.673 & \cellcolor{red!20}1.284 & 0.68 & 0.379 & \cellcolor{green!20}0.311 & 0.348  \\ \hline
                \texttt{1024$\times$768} & 1.627 & \cellcolor{red!20}2.975 & 1.551 & 0.833 & \cellcolor{green!20}0.626 & 0.628   \\ \hline
                \texttt{1920$\times$1200} & 4.65 & \cellcolor{red!20}8.796 & 4.499 & 2.362 & 1.513 & \cellcolor{green!20}1.224   \\ \hline
                \texttt{3840$\times$2160} & 21.556 & \cellcolor{red!20}36.28 & 18.509 & 9.594 & 5.533 & \cellcolor{green!20}3.732   \\ \hline
                \texttt{7680$\times$4320} & 92.397 & \cellcolor{red!20}156.381 & 78.988 & 40.655 & 22.315 & \cellcolor{green!20}13.553  \\ \hline
            \end{tabular}
            \caption{Average total time for different number of threads.}
        \end{table}
    \end{minipage}
\end{figure}

The use of include guards for this library allows us to truly see the difference between parallelized and sequential program that uses no library overhead and the difference is clearly seen in the first two columns. For small images 8 threads performed better than 16. The implementation using 16 threads was on average $3.759$-times faster than sequential one. For the largest image it was $6.817$-times faster.

\section{Dynamic Thread Optimization}

Fix this data analysis to be on 16 cores not 32 :/

We ran the algorithm using only one repetition allowing 16 threads, however setting each of the constants to different values. The \textbf{best values for the constants and time} can be seen below. This optimization managed to speed-up most image processing, especially for larger images. For smaller ones there was barely any noticable difference.

\begin{figure}[H]
    \begin{minipage}{\textwidth}
        \begin{table}[H]
            \centering
            \begin{tabular}{|l|p{1.8cm}|p{1cm}|p{1cm}|p{1.1cm}|} \hline
                \cellcolor{blue!20}\textbf{image} & \cellcolor{blue!20}\texttt{time (s)} & \cellcolor{blue!20}\texttt{C1} & \cellcolor{blue!20}\texttt{C2} & \cellcolor{blue!20}\texttt{C3} \\ \hline
                \texttt{720$\times$480} & \cellcolor{green!20}0.333 & 1000 & 500 & 100 \\ \hline
                \texttt{1024$\times$768} & \cellcolor{green!20}0.471 & 500 & 1000 & 100 \\ \hline
                \texttt{1920$\times$1200} & \cellcolor{green!20}0.778 & 1000 & 500 & 100 \\ \hline
                \texttt{3840$\times$2160} & \cellcolor{green!20}2.502 & 500 & 1000 & 100 \\ \hline
                \texttt{7680$\times$4320} & \cellcolor{green!20}9.154 & 500 & 1000 & 100 \\ \hline
            \end{tabular}
            \caption{Best heuristic per image, one run per configuration.}
        \end{table}
    \end{minipage}
\end{figure}

It is interasting to see that for all images a small constat was prefered when shifting pixles, meaning most if not all threads were utilized, therefore memory constraints did not impose as much overhead as expected. ...

\section{Greedy Seam Removal}

We ran the algorithm using different number of threads as seen in the first section however this also conditions the number of seams removed in one batch. Below are the times and the pictures obtained.

\begin{figure}[!hbpt]
    \begin{minipage}{\textwidth}
        \begin{table}[H]
            \centering
            \begin{tabular}{|l|*{5}{p{1cm}|}} \hline
                \cellcolor{blue!20}\textbf{image / threads} & \cellcolor{blue!20}\texttt{2} & \cellcolor{blue!20}\texttt{4} & \cellcolor{blue!20}\texttt{8} & \cellcolor{blue!20}\texttt{16} & \cellcolor{blue!20}\texttt{32} \\ \hline
                \texttt{720$\times$480}  & 0 & 0 & 0 & 0 & 0 \\ \hline
                \texttt{1024$\times$768} & 0 & 0 & 0 & 0 & 0 \\ \hline
                \texttt{1920$\times$1200}& 0 & 0 & 0 & 0 & 0 \\ \hline
                \texttt{3840$\times$2160}& 0 & 0 & 0 & 0 & 0 \\ \hline
                \texttt{7680$\times$4320}& 0 & 0 & 0 & 0 & 0 \\ \hline
            \end{tabular}
            \caption{Average total time for different implementations (placeholder).}
        \end{table}
    \end{minipage}
\end{figure}

\end{document}